---
title: "brooks lake data test"
author: "Samantha Peña"
date: "2024-11-22"
output: html_document
---

```{r}
# Install required packages (if not already installed)
install.packages("tidyverse")
install.packages("readxl")

# Load the libraries
library(tidyverse)  # For data manipulation and visualization
library(readxl)     # For reading Excel files

```


```{r}
# Specify the correct file path
file_path <- "~/Desktop/LAKE DATA/BROOKS LAKE DATA/WYDEQ 2021 Lander Phytoplankton Flat Data.xlsx"

# Read the Excel file and specify the correct sheet
data <- read_excel(path = file_path, sheet = "2021 Lander Phytoplankton")

# Check the first few rows of your dataset
head(data)

# Check the structure of your dataset
str(data)

```

.

```{r}
# Load stringr (part of tidyverse)
library(stringr)

# Filter rows where Sample_Station_Name contains "BROOKS"
brooks_data <- cyanobacteria_data %>%
  filter(str_detect(Sample_Station_Name, "Brooks Lake"))

# View the filtered data
head(brooks_data)

```

(a) Spatial Summary

Summarize the total abundance of cyanobacteria by location within the BROOKS dataset.
```{r}
# Summarize total abundance at BROOKS stations
brooks_spatial_summary <- brooks_data %>%
  group_by(Sample_Station_Name) %>%
  summarise(total_abundance = sum(TotalCells_L, na.rm = TRUE)) %>%
  arrange(desc(total_abundance))

# View the spatial summary
brooks_spatial_summary

```

Create a bar plot of cyanobacteria abundance at BROOKS stations.

```{r}
ggplot(brooks_spatial_summary, aes(x = reorder(Sample_Station_Name, total_abundance), y = total_abundance)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Cyanobacteria Abundance at Brooks ",
    x = "Station Name",
    y = "Total Abundance (cells/L)"
  ) +
  theme_minimal()

```
(a) Temporal Summary

Summarize cyanobacteria abundance over time at BROOKS stations.

```{r}
# Summarize abundance by date
brooks_temporal_summary <- brooks_data %>%
  group_by(Sample_Date_Collected) %>%
  summarise(total_abundance = sum(TotalCells_L, na.rm = TRUE)) %>%
  arrange(Sample_Date_Collected)

# View the temporal summary
brooks_temporal_summary

```
b) Visualize Temporal Trends

Create a line plot showing how cyanobacteria abundance changes over time at BROOKS stations.

```{r}
ggplot(brooks_temporal_summary, aes(x = Sample_Date_Collected, y = total_abundance, group = 1)) +
  geom_line(color = "blue", linewidth = 1) +
  labs(
    title = "Temporal Trends in Cyanobacteria Abundance at BROOKS Stations",
    x = "Date",
    y = "Total Abundance (cells/L)"
  ) +
  theme_minimal()

```
(a) Identify Dominant Species

Summarize cyanobacteria species abundance at BROOKS stations.

```{r}
# Summarize species abundance at BROOKS stations
brooks_species_summary <- brooks_data %>%
  group_by(Taxon) %>%
  summarise(total_abundance = sum(TotalCells_L, na.rm = TRUE)) %>%
  arrange(desc(total_abundance))

# View the dominant species
brooks_species_summary

```
b) Visualize Species-Specific Temporal Trends

Examine how individual species vary over time at BROOKS stations.

```{r}
ggplot(brooks_data, aes(x = Sample_Date_Collected, y = TotalCells_L, color = Taxon)) +
  geom_line(size = 1) +
  labs(
    title = "Temporal Trends of Cyanobacteria Species at BROOKS Stations",
    x = "Date",
    y = "Abundance (cells/L)",
    color = "Species"
  ) +
  theme_minimal()

```
##  Line Graphs (Temporal Trends of Individual Species)
```{r}
ggplot(brooks_data, aes(x = Sample_Date_Collected, y = TotalCells_L, color = Taxon, group = Taxon)) +
  geom_line(size = 1) +
  labs(
    title = "Temporal Trends of Cyanobacteria Species at BROOKS Stations",
    x = "Date",
    y = "Abundance (cells/L)",
    color = "Species"
  ) +
  theme_minimal()

```

## Stacked Area Chart (Total Composition Over Time)
```{r}
ggplot(brooks_data, aes(x = Sample_Date_Collected, y = TotalCells_L, fill = Taxon)) +
  geom_area(alpha = 0.8) +
  labs(
    title = "Relative Abundance of Cyanobacteria Species Over Time",
    x = "Date",
    y = "Abundance (cells/L)",
    fill = "Species"
  ) +
  theme_minimal()

```

## 3. Bar Plot (Dominant Species on Specific Dates)
```{r}
ggplot(brooks_data, aes(x = Sample_Date_Collected, y = TotalCells_L, fill = Taxon)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Dominant Cyanobacteria Species on Sampling Dates",
    x = "Date",
    y = "Abundance (cells/L)",
    fill = "Species"
  ) +
  theme_minimal()

```
##Heatmap (Species Abundance Across Time)

```{r}

install.packages("viridis")
library(viridis)

ggplot(brooks_data, aes(x = Sample_Date_Collected, y = Taxon, fill = TotalCells_L)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "Heatmap of Cyanobacteria Abundance by Species",
    x = "Date",
    y = "Species",
    fill = "Abundance (cells/L)"
  ) +
  theme_minimal()

```
##Boxplot (Species Distribution Across Stations or Time)

```{r}
ggplot(brooks_data, aes(x = Taxon, y = TotalCells_L, fill = Taxon)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Cyanobacteria Abundance by Species",
    x = "Species",
    y = "Abundance (cells/L)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
##Add in other years
# For Excel files
library(readxl)
new_data <- read_excel("path_to_new_file.xlsx")

# For CSV files
c"path_to_new_file.csv")



```{r}
library(readxl)
new_data <- read_excel("Desktop/LAKE DATA/BROOKS LAKE DATA/2020 Brooks Lake Phytoplankton Raw Data.xlsx")

```

```{r}
colnames(data)        # Existing data column names
colnames(new_data)    # New data column names

```

```{r}
# Combine the two datasets
combined_data <- bind_rows(data, new_data)

# Check the combined data
head(combined_data)

```

```{r}
# Save as CSV
write_csv(combined_data, "combined_cyanobacteria_data.csv")

# Save as Excel
install.packages("writexl")
library(writexl)
write_xlsx(combined_data, "combined_cyanobacteria_data.xlsx")

```


```{r}
# Install and load lubridate if not already installed
install.packages("lubridate")
library(lubridate)

# Add a Year column by extracting it from Sample_Date_Collected
new_data <- new_data %>%
  mutate(Year = year(Sample_Date_Collected))

# Preview the updated data
head(new_data)

```








```{r}
combined_data %>%
  group_by(Year, Sample_Date_Collected) %>%
  summarise(total_abundance = sum(TotalCells_L, na.rm = TRUE)) %>%
  ggplot(aes(x = Sample_Date_Collected, y = total_abundance, color = as.factor(Year))) +
  geom_line(size = 1) +
  labs(
    title = "Cyanobacteria Abundance Across Years",
    x = "Date",
    y = "Abundance (cells/L)",
    color = "Year"
  ) +
  theme_minimal()

```















